What is an Ancestral Recombination Graph?

### 2022-02-01 paper.tex
Tweaks

The state-space of the ARG is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.

@@

\section*{The ARG as a data structure}
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the miniumum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

@@

The first method to explicitly produce an ARG as output is SHRUB developed
by \citet{song2005efficient}.
% AI: I think \citep{song2003parsimonious} came before.

@@

This ambiguity can be dealt with in a number of ways. In the original
treatments, the direction in which ancestral material travels at
a breakpoint is handled by a simple rule, sites to the left of the
breakpoint go to the left parent and sites to right go to the
right parent. [Illustrate the problem with simulating an ARG
like this]. ~\cite{ignatieva2021kwarg} resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event. This is the same approach
as used in beagle.

>>>>

The state-space of the ARG is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

@@

\section*{The ARG as a data structure}
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the miniumum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

@@

@@

This ambiguity can be dealt with in a number of ways. In the original
treatments, the direction in which ancestral material travels at
a breakpoint is handled by a simple rule, sites to the left of the
breakpoint go to the left parent and sites to right go to the
right parent. [Illustrate the problem with simulating an ARG
like this]. ~\cite{ignatieva2021kwarg} resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event. This is the same approach
as used in beagle~\citep{lyngso2005minimum}.
